---
title: "Template for an epidemiological analysis"
author: "Sanne Hoepel and Anna Streiber"
date: "Last edited on `r format(Sys.time(), '%d-%m-%Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

<!-- The code below specifies which code- and output chunks are scrollable.
CSS stands for cascading style sheet and is used to change the appearance of an html page.
You can remove this if you DO NOT want scrollable code chunks in your HTML -->


```{css, echo=FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{css, echo=FALSE}
.scroll-300 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, ft.htmlscroll = F)
```

# 1. Analysis Plan

## 1.1 Introduction

*Here you write the background for your project. This is the main part
of your project proposal.*

This can be written in bullet points or text.


## 1.2 Hypothesis

It's always good to write out your research question and associated
hypotheses explicitly. This makes it easier for you to determine which
analysis you have to use, but it also improves understanding for the
reader/ researcher who wants to understand (and maybe replicate) your
analysis.


## 1.3 Methods

In this section, you can write the methods of the project in bullet points. The more
work you put into this now, the less time you need to invest into the
methods section of your manuscript. We have added some suggested headings you could use. 

### 1.3.1 Study setting and population



### 1.3.2 Exposure



### 1.3.3 Outcome



### 1.3.4 Covariates



### 1.3.5 Statistical analyses


## 2. Set up R

### 2.1 Packages

When you load your packages, please make sure to have the `here` package installed. It allows to build to-level directory paths that are less fragile than the `setwd` function. If you specify your paths well, it will aid in replicating your results if you share you analysis folder.

```{r }
## Load packages 
library(package_name)   # here, you can write a short explanation of what the package does

library(here)           # top-level directory to build paths
```

### 2.2 Version control

Below, you see the code that specifies which R, R studio, and package versions you used for this project.

```{r, echo=FALSE}
getRInfo <- function() {
  # Get R version
  r_version <- R.version

  # Get RStudio version
  rstudio_info <- rstudioapi::versionInfo()
  rstudio_version <- paste("RStudio Version:", rstudio_info$version, "(", rstudio_info$mode, ")", sep = " ")
  
  cat(r_version$version.string, "\n")
  cat(rstudio_version, "\n")
}


# Call the function to display the information
getRInfo()

# List of packages
packages <- c("package_name", "here")

# Print package versions
for (pkg in packages) {
  version <- as.character(packageVersion(pkg))
  cat(pkg, "version :", version, "\n")
}
```

## 3. Load data

The data for the project needs to loaded from different sources,
compiled into one dataset.

Data can be downloaded from the epidemiology data
[Wiki](https://epi-wiki.erasmusmc.nl/wiki/). Please refer to our tips
and tricks html document, for a detailed overview on how to download
your data.

```{r}
example <- read.csv(here("your_directory/folder/yourdata.csv"))
```

### 3.1 Exposure
Load data on your exposure and (if necessary) recode

```{r}

```

### 3.2 Outcome
Load data on your outcome and (if necessary) recode

```{r}

```

### 3.3 Covariates
Load data on your covariates and (if necessary) recode

```{r}

```

### 3.4 Final Dataset

In a next step, we only select the variables that we will need in the
analysis.

```{r}


```

We select the participants that are included in our analysis.

```{r}

```

# 4. Data Exploration

## 4.1 Study population characteristics {.tabset}

The data distributions of the determinant, the outcome, and the
covariates can be visualized using bar graphs and histograms. Refer
to [this chapter](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) of
the R cookbook to change the colors of your graphs.

Afterward, you can print sample characteristics in Table 1.

### Create figures

```{r}

```

```{r, fig.show='hide'}

```

### View figures

```{r, echo=FALSE}

```

```{r, echo=FALSE}

```

### Create table 1

Tip: Label all variables to make the Table 1 result easier to interpret.


```{r}

```

You may want to consider to save your code in an Excel table. This can
be done using the `write.xlsx` function.

```{r}

```

### View table 1

```{r, echo=FALSE}

```

## 4.2 Imputation

If you have missing data, you can perform (multiple) imputation in this
section. The mice algorithm (van Buuren and Groothuis-Oudshoorn 2011) is
the most common way to impute data in R.

Check the script repository for example scripts.

# 5. Hypothesis Testing {.tabset}

## 5.1 Analyses

```{r}

```

Save the results of the analysis.

```{r}

```

## 5.2 Visualized results {.tabset}

Here, you can visualize the results from the main analyses, for example using tables and graphs. 

```{r}

```

# 6. Sensitivity Analyses {.tabset}

For example: 

- Stratification for cohort
- Do you see differential effects?
- Are there interactions?

# 7. Data Interpretation
Write down your interpretation of your results. This makes it easy to discuss with your collaborators. 

# 8. Logbook {.tabset}

*THESE SECTIONS CAN BE REPEATED DEPENDING ON NUMBER OF MEETINGS*

## Explanation
We recommend including this little explanation section in your Markdown. It helps other people to understand what you have done and highlights the importance of transparency in research.

In this section, we provide a little logbook of meetings that we had concerning this research project. The idea is, to increase transparency of decisions that were made during the research project regarding data analysis and data interpretation.

## Meetings

Meeting 1 (dd/mm/yyyy)

:   Summary of meeting 1 (incl. people present)

Meeting 2 (dd/mm/yyyy)

:   Summary of meeting 2 (incl. people present)

# 9. Manuscript {.tabset}

## Analyses

Here, you can provide an overview of the analysis that you would like to
include in your paper
```{r}

```

## Tables

Tables that will be included in your manuscript
```{r}

```

## Figures

Figures that will be included in your manuscript
```{r}

```

