---
title: "Smoking and mortality"
author: "Sanne Hoepel and Anna Streiber"
date: "Last edited on `r format(Sys.time(), '%d-%m-%Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, ft.htmlscroll = F)
```

# 1. Analysis Plan

## 1.1 Introduction

*Here you write the background for your project. This is the main part
of your project proposal.*

Smoking has been identified as a global public health concern,
attributed to its substantial association with various adverse health
outcomes, including cancer, cardiovascular diseases, and respiratory
conditions. Among these, mortality stands as a critical endpoint,
illustrating the ultimate consequence of smoking on individuals'
well-being. Understanding the relationship between smoking and mortality
is paramount for public health strategies, policy formulation, and
personalized healthcare interventions.

In this study, we employ survival analysis to investigate the
relationship between smoking habits and mortality risk. Survival
analysis, a statistical methodology well-suited for time-to-event data,
allows for a comprehensive assessment of mortality patterns in a
population over time, while accounting for censoring and other potential
confounding variables. By utilizing this approach, we aim to shed light
on the magnitude of mortality risk associated with smoking, discerning
the impact of smoking intensity, duration, and cessation on survival
probabilities.

By examining observational data, this study aims to contribute to the
growing body of evidence regarding smoking-related mortality, providing
valuable insights to guide public health policies and individual
choices. A deeper comprehension of the risks associated with smoking and
its influence on mortality is crucial for developing effective
strategies to mitigate the adverse effects and improve overall health
outcomes.

## 1.2 Hypothesis

It's always good to write out your research question and associated
hypotheses explicitly. This makes it easier for you to determine which
analysis you have to use, but it also improves understanding for the
reader/ researcher who wants to understand (and maybe replicate) your
analysis.

In this example, our research question is: To what extent is smoking
related to mortality?

An explicit hypothesis is that current smokers (*exposure*) have a
higher risk (*directionality of effect*) of mortality (*outcome*).

## 1.3 Methods

In this section, you can write the methods of the project in bullet points. The more
work you put into this now, the less time you need to invest into the
methods section of your manuscript.

### 1.3.1 Study setting and population

Here, you should explain where you got the data from. Think about who is
the source population and what is the study setting.

This study is embedded in the Rotterdam Study. The Rotterdam Study is a
prospective population-based cohort study that addresses determinants
and occurrence of cardiovascular, neurological, ophthalmologic,
psychiatric, and endocrine diseases in the elderly. At present the
Rotterdam Study incorporates three cohorts that were established in
1989, 2000, 2006, and 2015 respectively.

In 1989 all residents of Ommoord, a suburb of Rotterdam, aged 55 years
and over were invited to participate. A total of 7,983 out of 10,275 men
and women entered the study (response rate 78 percent). Baseline data
were collected from 1990 until 1993. After that, participants had
regular follow-up visits scheduled every 3 to 6 years.

### 1.3.2 Exposure: smoking

Smoking status as been assessed during an extensive home interview.
Based on the interview, participants were categorized into current,
former, and never smokers. Cigarette-, cigar-, and pipe smoking have
been considered.

### 1.3.3 Outcome: mortality

In this study, all-cause mortality served as the outcome. Mortality was
obtained by:

-   notification by the municipal administration
-   notification by the GP or nursing home
-   notification by the family

### 1.3.4 Covariates

The covariates of interest included age in years and sex (male/female).

### 1.3.5 Statistical analyses

We used Cox proportional hazards models to examine the association of
smoking with incident mortality. Hazard ratios (HR) and 95% confidence
intervals (CI) were calculated. All analyses were
performed using R version 4.3.0. (Team 2023) 

## 2. Set up R

### 2.1 Version control

Below, you can see the R and R studio versions that have been used in this project.

```{r, echo=FALSE}
getRInfo <- function() {
  # Get R version
  r_version <- R.version

  # Get RStudio version
  rstudio_info <- rstudioapi::versionInfo()
  rstudio_version <- paste("RStudio Version:", rstudio_info$version, "(", rstudio_info$mode, ")", sep = " ")
  
  cat(r_version$version.string, "\n")
  cat(rstudio_version, "\n")
}


# Call the function to display the information
getRInfo()
```

### 2.2 Packages

```{r }
## Load packages 
library(rstudioapi)   # version control
library(haven)        # Read/write various data formats.
library(tidyr)        # Tidy and restructure data.
library(dplyr)        # Data manipulation and transformation.
library(here)         # Top-level directory to build paths.
library(ggplot2)      # Data visualization.
library(xlsx)         # Read/write Excel files.
library(openxlsx)     # Read/write Excel files.
library(table1)       # Create descriptive tables.
library(mice)         # Multiple imputation of missing data.
library(kableExtra)   # Create elegant tables.
library(RColorBrewer) # Change colors of your plot.
library(survival)     # Performing survival analysis.
```

## 3. Load data

The data for the project needs to loaded from different sources,
compiled into one dataset.

Data can be downloaded from the epidemiology data
[Wiki](https://epi-wiki.erasmusmc.nl/wiki/). Please refer to our tips
and tricks html document, for a detailed overview on how to download
your data.

```{r}
# Datasets are loaded into R using the read_spss function
basics <- read_spss(here("V:/Uitwissel/Anna_Streiber/Anna & Sanne/R Markdown template/Data/RoterdamStudy_Basics2014.sav"))

smoking_e1 <- read_spss(here("V:/Uitwissel/Anna_Streiber/Anna & Sanne/R Markdown template/Data/e1_intvw_SMOKING_(23-nov-2011).sav"))

mortality_data  <- read_spss(here("V:/Uitwissel/Anna_Streiber/Anna & Sanne/R Markdown template/Data/fp_VitalStatus_2022-42.sav"))

response_data <- read_spss(here("V:/Uitwissel/Anna_Streiber/Anna & Sanne/R Markdown template/Data/Ergo1ResponseDetail_(22-jan-2015)_excerpt.sav"))

# Datasets are merged using the full_join function 
dataset <- 
  full_join(basics, smoking_e1, by = 'ergoid') %>%
  full_join(mortality_data, by = 'ergoid') %>%
  full_join(response_data, by = 'ergoid')
```

### 3.1 Exposure: smoking

Create a 'current', 'former', 'never' variable for smoking:

-   0 = never
-   1 = former
-   2 = current, NA

Please refer to our [data cleaning
guide](https://rs-data-guide.netlify.app/) for more detailed
information.

```{r}
## Categorize smoking 
dataset <- dataset %>%
mutate(
    smoke = ifelse((e1_ai7_20 == 0 & e1_ai7_30 == 0), 0, NA),
    smoke = ifelse((e1_ai7_20 == 0 & e1_ai7_30 == 1), 1, smoke),
    smoke = ifelse(e1_ai7_20 == 1, 2, smoke), 
    smoke = factor(smoke, levels = c(0:2), labels = c("never", "former", "current")))

```

### 3.2 Outcome: Mortality

The mortality data contains a variable called 'fp_vitalstatus' that
states whether someone was still alive at the latest contact.

The variable 'fp_censordate' is the variable containing the date of last
contact.

The code below integrates the mortality variable into the dataset and
calculates the follow-up time in years (related to mortality analysis)
based on the difference between the date of the last contact
(fp_censordate) and the baseline visit (baseline_date).

```{r}
dataset <- dataset %>%
  
      ## rename the mortality variable and baseline date for clarity 
    rename(mortality = fp_vitalstatus, 
           baseline_date = e1_aintdat
           )  %>%
  
    ## calculate the follow-up time in years 
    mutate(followup_mortality = as.numeric(difftime(fp_censordate, baseline_date, units = 'weeks')/52.25))

```

### 3.3 Covariates: Age and sex

Age is calculated based on the difference between the date of birth and:

-   the last center visit
-   loss to follow-up
-   death or 01-01-2020

**whichever occurs first.**

```{r}
dataset <- dataset %>%
          ## save sex as a factor
  mutate(sex = factor(sex, levels = c(0,1), labels = c("male", "female")), 
           ##calculate the age at baseline
           age_baseline  = as.numeric(difftime
                                      (baseline_date, date_of_birth, units = "weeks")/52.25))
```

### 3.4 Final Dataset

In a next step, we only select the variables that we will need in the
analysis.

```{r}
dataset <- dataset %>%
  select(ergoid, smoke, mortality, followup_mortality, sex, age_baseline)

```

We select the participants that are included in our analysis. Our
inclusion requirements are: complete data on smoking at baseline and
follow-time for mortality (implying they gave informed consent for
follow-up).

**Note. This can be different for a different project!**

```{r}
dataset <- dataset %>%
  #select the participants with complete data on smoking at baseline
  filter(!is.na(smoke) & followup_mortality > 0)
```

# 4. Data Exploration

## 4.1 Study population characteristics {.tabset}

The data distributions of the determinant, the outcome, and the
covariates have been visualized using bar graphs and histograms. Refer
to [this chapter](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) of
the R cookbook to change the colors of your graphs.

Afterward, all sample characteristics have been printed in table 1.

### Create figures

```{r}
# Bargraph to show smoking and mortality
smoke_mort <- ggplot(dataset, aes(x = mortality, fill = smoke)) +
  geom_bar() +
  labs(title = "Mortality by Smoking Status",
       x = "Mortality",
       y = "Count") +
  facet_wrap(~smoke, ncol = 1) +
  scale_fill_manual(values = c("never" = "#CC6666", "former" = "#9999CC", "current" = "#66CC99"))
```

```{r, fig.show='hide'}
# Histogram to show the distribution of age
age <- ggplot(dataset, aes(x = age_baseline)) +
  geom_histogram(binwidth = 5, fill = "#009E73", color = "white") +
  labs(title = "Age Distribution",
       x = "Age (years)",
       y = "Count")
```

### View figures

```{r, echo=FALSE}
age
```

```{r, echo=FALSE}
smoke_mort
```

### Create table 1

Label all variables to make the Table 1 result easier to interpret.

When creating the table, use `as.factor` for the parameters containing
the events. They need to stay numeric for the `coxph` function.

```{r}
label(dataset$age_baseline) <- 'Age at baseline'
label(dataset$mortality) <- 'Follow up time'
label(dataset$followup_mortality) <- 'Follow up time'
label(dataset$sex) <- 'Sex'
label(dataset$smoke) <- 'Smoking'


#Create table 1 using the table 1 function 
table1 <- as.data.frame(table1(~ age_baseline + followup_mortality + sex 
                                   + smoke +  as.factor(mortality) ,  data = dataset))
```

You may want to consider to save your code in an Excel table. This can
be done using the `write.xlsx` function.

```{r, eval = FALSE}
# Save the table 1 to excel
write.xlsx(table1, "Insert your directory here")
```

### View table 1

```{r, echo=FALSE}
table1
```

## 4.2 Imputation

If you have missing data, you can perform (multiple) imputation in this
section. The mice algorithm (van Buuren and Groothuis-Oudshoorn 2011) is
the most common way to impute data in R.

Check the script repository for example scripts.

# 5. Hypothesis Testing {.tabset}

## 5.1 Smoking \~ mortality

```{r}
model1 <- coxph(Surv(followup_mortality, mortality) ~ smoke + age_baseline + sex, data = dataset )

summary(model1)
```

Save the results of the analysis.

```{r}
table_results <- data.frame(matrix(NA, 2, 4)) 

model1_sum <- summary(model1)

table_results[1, 1:4] <- data.frame(round(model1_sum$coefficients[1,2],2), 
                              round(model1_sum$conf.int[1,3],2),
                              round(model1_sum$conf.int[1,4],2),
                              round(model1_sum$coefficients[1,5], 3))
table_results[2, 1:4] <- data.frame(round(model1_sum$coefficients[2,2],2), 
                              round(model1_sum$conf.int[2,3],2),
                              round(model1_sum$conf.int[2,4],2),
                              round(model1_sum$coefficients[2,5], 3))

# paste confidence intervals 
  for(i in 1 : nrow(table_results)){ 
    table_results[i,2] <- (paste(table_results[i,2], table_results[i,3], sep = '; '))
  }
  table_results <- table_results[ , c(1,2,4)]
  
  names(table_results) <- c('Hazard Ratio', '95% CI', 'p') 
  
  rownames(table_results) <- c("former smoker", "current smoker")
```

## 5.2 Visualized results {.tabset}

Here, you can visualize the results from the main analyses.

### Smoking and mortality

```{r}
kable(table_results)
```

# 6. Sensitivity Analyses {.tabset}

Stratification for cohort \> do you see different effects \> do formal
interaction .

# 7. Data Interpretation

# 8. Logbook {.tabset}

*THESE SECTIONS CAN BE REPEATED DEPENDING ON NUMBER OF MEETINGS*

## Explanation
In this section, we provide a little logbook of meetings that we had concerning this research project. The idea is, to increase transparency of decisions that were made during the research project regarding data analysis and data interpretation.

## Meetings

Meeting 1 (dd/mm/yyyy)

:   Summary of meeting 1 (incl. people present)

Meeting 2 (dd/mm/yyyy)

:   Summary of meeting 2 (incl. people present)

# 9. Manuscript {.tabset}

## Analyses

Here, you can provide an overview of the analysis that you would like to
include in your paper

## Tables

Tables that will be included in your manuscript

## Figures

Figures that will be included in your manuscript
